---
title: 'HCA-DDI Model'
author: "Meenakshi Srinivasan"
output:
  pdf_document: default
  html_document: default
---
## Model description for Strategy 1 pill only- 
- This model does not consider a post-imb state and assumes that women can only enter the imb state once and continue in that state (constantly).
- Does not consider age specific probabilities for birth, ia, sa, and ep

## Need to do univariate sensitivity analysis 
eg 1. p_vte 0.0000833326 (0.00005-0.00010)

```{r, eval = FALSE,echo=FALSE}
Developed by the Decision Analysis in R for Technologies in Health (DARTH) workgroup:

Fernando Alarid-Escudero, PhD (1) 

Eva A. Enns, MS, PhD (2)	

M.G. Myriam Hunink, MD, PhD (3,4)

Hawre J. Jalal, MD, PhD (5) 

Eline M. Krijkamp, MSc (3)	

Petros Pechlivanoglou, PhD (6,7)

Alan Yang, MSc (7)

In collaboration of: 		

1. Drug Policy Program, Center for Research and Teaching in Economics (CIDE) - CONACyT, 
   Aguascalientes, Mexico
2. University of Minnesota School of Public Health, Minneapolis, MN, USA
3. Erasmus MC, Rotterdam, The Netherlands
4. Harvard T.H. Chan School of Public Health, Boston, USA
5. University of Pittsburgh Graduate School of Public Health, Pittsburgh, PA, USA
6. University of Toronto, Toronto ON, Canada
7. The Hospital for Sick Children, Toronto ON, Canada

Please cite our publications when using this code:
 
- Jalal H, Pechlivanoglou P, Krijkamp E, Alarid-Escudero F, Enns E, Hunink MG. 
An Overview of R in Health Decision Sciences. Med Decis Making. 2017; 37(3): 735-746. 
https://journals.sagepub.com/doi/abs/10.1177/0272989X16686559
 
- Krijkamp EM, Alarid-Escudero F, Enns EA, Jalal HJ, Hunink MGM, Pechlivanoglou P. 
Microsimulation modeling for health decision sciences using R: A tutorial. 
Med Decis Making. 2018;38(3):400–22. 
https://journals.sagepub.com/doi/abs/10.1177/0272989X18754513
 
- Krijkamp EM, Alarid-Escudero F, Enns E, Pechlivanoglou P, Hunink MM, Jalal H. 
A Multidimensional Array Representation of State-Transition Model Dynamics. 
Med Decis Making. Online First https://doi.org/10.1177/0272989X19893973

Copyright 2017, THE HOSPITAL FOR SICK CHILDREN AND THE COLLABORATING INSTITUTIONS. 
All rights reserved in Canada, the United States and worldwide. Copyright, 
trademarks, trade names and any and all associated intellectual property are 
exclusively owned by THE HOSPITAL FOR Sick CHILDREN and the collaborating 
institutions. These materials may be used, reproduced, modified, distributed 
and adapted with proper attribution.
```


```{r}
rm(list = ls())      # clear memory (removes all the variables from the workspace)
```

# 01 Load packages

```{r}
if (!require('pacman')) install.packages('pacman'); library(pacman) # use this package to conveniently install other packages
# load (install if required) packages from CRAN
p_load("diagram", "tidyverse", "here") 
```

# 02 Load functions

```{r}
ProbToRate <- function(p){
  # argument
  # p : the probability
  # Retunrs:
  # r : rate
  r <- -log(1 - p)
  r
}

RateToProb <- function(r, t){
  p <- 1 - exp(-r * t)
  return(p)
}
```

# 03 Input model parameters

```{r}
# Strategy names
v_names_str <- c("pill-only", "pill-plus-inducer", "pill-plus-inhibitor")  

# Number of strategies
n_str <- length(v_names_str)

# Markov model parameters
# state names- 9 states
v_n  <- c("initial", "birth", "induced-abortion",         
          "spontanous-abortion", "ectopic-pregnancy",
          "vte", "mi", "stroke", "imb")    

# number of states
n_states  <- length(v_n)                                  
                                
v_age <- c(15:44)  #vector of ages

n_t <- 12 # 12 cycles of 1 month duration

# Monthly probability of failure of pill-only, pill-plus-inducer and pill-plus-inhibitor
p_failure_pill <- 0.0009245723
p_failure_pill_inducer <- 0.001299155
p_failure_pill_inhibitor <- 0.0006830999

# Monthly probability of birth
p_birth <- 0.03730147
p_ia <- 0.04938104
p_sa <- 0.01523964
p_ep <- 0.0008371774

# Probability of unintended pregnancy outcomes by strategy
# Strategy 1 pill-only
p_birth_pill <- p_birth * p_failure_pill
p_ia_pill <- p_ia * p_failure_pill
p_sa_pill <- p_sa * p_failure_pill
p_ep_pill <- p_ep * p_failure_pill

# Probability of adverse events - general
p_vte <- 0.0000833326
p_mi <- 0.00001814990
p_stroke <- 0.00003085469

# Probability of serious adverse events - pill-only
p_vte_pill <- p_vte
p_mi_pill <- p_mi
p_stroke_pill <- p_stroke

### Probability of IMB - pill-only ###

### Probability of IMB - pill-plus-inhibitor ### 
# Note: These probabilities will be assumed to be the same as for the pill-alone group

# Monthly probability for IMB
# 12 month probability = 0.5% = 0.005
# 1 month probability

ProbToRate(0.005)
RateToProb(0.005012542, 1/12)

p_imb_pill <- 0.0004176246     # % of women with IMB

# Costs of initial state for three strategies

c_pill           <- 52          # cost of remaining in pill only group

# Cost of birth outcomes 
c_birth  <- 13252              # cost of remaining one cycle sick
c_ia     <- 931                # cost of remaining one cycle dead
c_sa     <- 1141               # cost of spontaneous abortion
c_ep     <- 6174               # cost of ectopic pregnancy

# cost of adverse event outcomes
c_vte     <- 14305                # cost of venous thromboembolism
c_mi      <- 23528                # cost of myocardial infarction
c_stroke  <- 28979                # cost of stroke

c_imb     <- 50                   # cost of intermenstrual bleeding (conservative)
 

# Utility values 

## Utility of initial state for three strategies 
# Assuming perfect health on pill alone  
u_pill           <-  1                       
# Lowered utility because of condition that requires drug (epilepsy, HIV etc)       
  
u_birth  <- 0.9064              # Utility of birth     
u_ia     <- 0.9615              # Utility of abortion
u_sa     <- 0.9423              # Utility of spontaneous abortion
u_ep     <- 0.91667             # Utility of ectopic pregnancy
  
u_vte    <- 0.81                # Utility of VTE
u_mi     <- 0.67                # Utility of MI
u_stroke <- 0.33                # Utility of Stroke
u_imb    <- 0.78                # Utility of Inter menstrual bleeding

```


# 04 Define and initialize matrices and vectors

## 04.1 Cohort trace for the strategy 1 - "Pill only"

```{r}
# create the cohort trace
m_M_pill <- matrix(NA, 
              nrow = n_t + 1 ,  # create Markov trace (n.t + 1 because R doesn't 
                                # understand Cycle 0)
              ncol = n_states, 
              dimnames = list(0:n_t, v_n))

m_M_pill[1, ] <- c(1, rep(0, n_states - 1))          # initialize first cycle of Markov trace
```

## 04.2 Transition probability matrix for the strategy 1 - "Pill only"

```{r}
# create the transition probability matrix
m_P_pill  <- matrix(0,
               nrow = n_states, ncol = n_states,
               dimnames = list(v_n, v_n)) # name the columns and rows of the transition 
                                          # probability matrix
m_P_pill
```

Fill in the transition probability matrix for the strategy 1 - "Pill only":

```{r}

  m_P_pill["initial", "initial" ]    <- 1 - p_birth_pill - p_ia_pill -
  p_sa_pill - p_ep_pill - p_vte_pill - p_mi_pill - p_stroke_pill - p_imb_pill 
  
  m_P_pill["initial", "birth" ]      <- p_birth_pill
  
  m_P_pill["initial", "induced-abortion" ]      <- p_ia_pill 
  m_P_pill["initial", "spontanous-abortion" ]  <- p_sa_pill  
  m_P_pill["initial", "ectopic-pregnancy" ]   <- p_ep_pill  
  m_P_pill["initial", "vte" ] <-  p_vte_pill
  m_P_pill["initial", "mi" ]  <- p_mi_pill
  m_P_pill["initial", "stroke" ] <-  p_stroke_pill 
  m_P_pill["initial", "imb" ]  <- p_imb_pill 
  m_P_pill["birth", "birth" ] <- 1
  m_P_pill["induced-abortion", "induced-abortion" ] <- 1
  m_P_pill["spontanous-abortion", "spontanous-abortion" ] <- 1
  m_P_pill["ectopic-pregnancy", "ectopic-pregnancy" ] <- 1
  m_P_pill["vte", "vte" ] <- 1
  m_P_pill["mi", "mi" ] <- 1
  m_P_pill["stroke", "stroke" ] <- 1
  m_P_pill["imb", "imb"] <- 1

# check rows add up to 1
rowSums(m_P_pill)
```

# 05 Run Markov model for the strategy 1 - "Pill only"

```{r}
for (t in 1:n_t){                   # loop through the number of cycles
  m_M_pill[t + 1, ] <- m_M_pill[t, ] %*% m_P_pill  # estimate the state vector for the next cycle (t + 1)
}

rowSums(m_M_pill)
```

# 06 Compute and Plot Outcomes for the strategy 1 - "Pill only"

## 06.1 Cohort trace for the strategy 1 - "Pill only"

```{r}
df_M_pill <- as.data.frame(m_M_pill) %>%
  tibble::rowid_to_column("cycle") %>%
  tidyr::gather(state, probability, -1)

ggplot(df_M_pill, aes(cycle, probability, color = state)) +
  geom_line()

```

## 06.2 Total Probability for Unintended pregnancy and adverse event outcomes for the strategy 1 - "Pill only"

```{r}
#preliminary results
v_up_pill <- rowSums(m_M_pill[, 2:5])  # calculating the UP probability   

v_adverseevents_pill <- rowSums(m_M_pill[, 6:9])
n_women <- 250000
n_births_pill <- m_M_pill[13, "birth"] * n_women
n_ia_pill <- m_M_pill[13, "induced-abortion"] * n_women
n_vte_pill <- m_M_pill[13, "vte"] * n_women
n_mi_pill <- m_M_pill[13, "mi"] * n_women
n_stroke_pill <- m_M_pill[13, "stroke"] * n_women
n_imb_pill <- m_M_pill[13, "imb"] * n_women
```


# 07 Compute Cost-Effectiveness Outcomes for the strategy 1 - "Pill only"

## 07.1 Mean Costs and QALYs for the strategy 1 - "Pill only"

```{r}
# per cycle
# calculate expected costs by multiplying m_M with the cost vector for the different 
# health states   
v_tc_pill <- m_M_pill %*% c(c_pill, c_birth, c_inducedabortion, c_spontaneousabortion,
                  c_ectopicpregnancy, c_vte, c_mi, c_stroke, c_imb)  
# calculate expected QALYs by multiplying m_M with the utilities for the different 
# health states   
v_tu_pill <- m_M_pill %*% c(u_pill, u_birth, u_inducedabortion, u_spontaneousabortion,
                  u_ectopicpregnancy, u_vte, u_mi, u_stroke, u_imb)  
```

## 07.2 Discounted Mean Costs and QALYs for the strategy 1 - "Pill only"

```{r}
# Discount costs by multiplying the cost vector with discount weights (v_dw) 
# v_tc_d_pill <-  t(v_tc_pill) %*% v_dwc
# Discount QALYS by multiplying the QALYs vector with discount weights (v_dw)
# v_te_d_pill <-  t(v_tu_pill) %*% v_dwe
```

## 07.3 Results for the strategy 1 - "Pill only"

```{r}
results_pill <- data.frame( "Total Discounted Cost" = v_tc_d_pill, 
                   
                       "Total Discounted QALYs" = v_te_d_pill, 
                       check.names = F)
results_pill
```
