---
title: 'HCA-DDI Model'
author: "Meenakshi Srinivasan"
output:
  pdf_document: default
  html_document: default
---
## Changes made-
1) Converted all probabilities to monthly probailities
2) Added 2 extra states

## Need to do-
1) Make transition array to account for changing probabilities of women who experience IMB
2) Accruing costs for women who experience IMB depending upon months
3) Interpretation of matrix. 


```{r, eval = FALSE,echo=FALSE}
Developed by the Decision Analysis in R for Technologies in Health (DARTH) workgroup:

Fernando Alarid-Escudero, PhD (1) 

Eva A. Enns, MS, PhD (2)	

M.G. Myriam Hunink, MD, PhD (3,4)

Hawre J. Jalal, MD, PhD (5) 

Eline M. Krijkamp, MSc (3)	

Petros Pechlivanoglou, PhD (6,7)

Alan Yang, MSc (7)

In collaboration of: 		

1. Drug Policy Program, Center for Research and Teaching in Economics (CIDE) - CONACyT, 
   Aguascalientes, Mexico
2. University of Minnesota School of Public Health, Minneapolis, MN, USA
3. Erasmus MC, Rotterdam, The Netherlands
4. Harvard T.H. Chan School of Public Health, Boston, USA
5. University of Pittsburgh Graduate School of Public Health, Pittsburgh, PA, USA
6. University of Toronto, Toronto ON, Canada
7. The Hospital for Sick Children, Toronto ON, Canada

Please cite our publications when using this code:
 
- Jalal H, Pechlivanoglou P, Krijkamp E, Alarid-Escudero F, Enns E, Hunink MG. 
An Overview of R in Health Decision Sciences. Med Decis Making. 2017; 37(3): 735-746. 
https://journals.sagepub.com/doi/abs/10.1177/0272989X16686559
 
- Krijkamp EM, Alarid-Escudero F, Enns EA, Jalal HJ, Hunink MGM, Pechlivanoglou P. 
Microsimulation modeling for health decision sciences using R: A tutorial. 
Med Decis Making. 2018;38(3):400–22. 
https://journals.sagepub.com/doi/abs/10.1177/0272989X18754513
 
- Krijkamp EM, Alarid-Escudero F, Enns E, Pechlivanoglou P, Hunink MM, Jalal H. 
A Multidimensional Array Representation of State-Transition Model Dynamics. 
Med Decis Making. Online First https://doi.org/10.1177/0272989X19893973

Copyright 2017, THE HOSPITAL FOR SICK CHILDREN AND THE COLLABORATING INSTITUTIONS. 
All rights reserved in Canada, the United States and worldwide. Copyright, 
trademarks, trade names and any and all associated intellectual property are 
exclusively owned by THE HOSPITAL FOR Sick CHILDREN and the collaborating 
institutions. These materials may be used, reproduced, modified, distributed 
and adapted with proper attribution.
```


```{r}
rm(list = ls())      # clear memory (removes all the variables from the workspace)
```

# 01 Load packages

```{r}
if (!require('pacman')) install.packages('pacman'); library(pacman) # use this package to conveniently install other packages
# load (install if required) packages from CRAN
p_load("diagram") 
```

# 02 Load functions

```{r}
# no functions required
```

# 03 Input model parameters

```{r}
# Strategy names
v_names_str <- c("Base Case", "Pill + Inducer", "Pill + Inhibitor")  

# Number of strategies
n_str <- length(v_names_str)

# Markov model parameters
v_n  <- c("Initial", "Birth", "Induced abortion",         # state names- 11 states
          "Spontanous abortion", "Ectopic pregnancy",
          "VTE", "MI", "Stroke", "IMB", "IMB quitters", "IMB continuers")    
n_states  <- length(v_n)                                  # number of states
                                

# monthly cycles
t_years             <- 1                # time horizon in years
t_cl_month          <- 1                # cycle length in months
n_CpY               <- 12/t_cl_month    # number of cycles per year
n_YpC               <- 1/n_CpY          # number of years per cycle
n_t                 <- (t_years * 12) / t_cl_month  # number of model cycles

# p = 1 - exp(-r*dt) # p = probability, r = rate, dt = period of time probability is defined for eg 1 year
# 
# r_up_pill = 1.11   # number of failures per 100 woman-years of exposure for 1 year
# r_up_pill = 0.0111 # number of failures per 1 woman-years of exposure for 1 year
# 
# p_yearly = 1 - exp(-0.0111) 
# 
# p_monthly = 1 - exp(-0.0111/12)

ProbToRate <- function(p){
  # argument
  # p : the probability
  # Retunrs:
  # r : rate
  r <- -log(1 - p)
  r
}
RateToProb <- function(r, t){
  p <- 1 - exp(-r * t)
  return(p)
}

# *Yearly* *rate* of unintended pregnancy (UP)- Contraceptive failure rate

rate_pill_up <- 0.0111             # rate of unintended pregnancy (UP) per 1 woman-year with pill alone
rate_pill_inducer_up <- 0.0156     # rate of UP per 1 woman-year with pill + enzyme inducer
rate_pill_inhibitor_up <- 0.0082   # rate of UP per 1 woman-year with pill + enzyme inhibitor

# *Monthly* *probability* of unintended pregnancy (UP)- Contraceptive failure probability
p_m_pill_up <- RateToProb(rate_pill_up, 1/12)
p_m_pill_inducer_up <- RateToProb(rate_pill_inducer_up, 1/12)
p_m_pill_inhibitor_up <- RateToProb(rate_pill_inhibitor_up, 1/12)

# *yearly* Conditional probability of Unintended pregnancy outcomes 
# These sum to 1

p_birth <- 0.3663               # probability of birth
p_inducedabortion <- 0.4554     # probability of induced abortion
p_spontaneousabortion <- 0.1683 # probability of spontaneous abortion
p_ectopicpregnancy <- 0.01      # probability of ectopic pregnancy

# *monthly* rate of Unintended pregnancy outcomes 
rate_m_birth <- ProbToRate(p_birth)
rate_m_inducedabortion <- ProbToRate(p_inducedabortion)/12
rate_m_spontaneousabortion <- ProbToRate(p_spontaneousabortion)/12
rate_m_ectopicpregnancy <- ProbToRate(p_ectopicpregnancy)/12

# *monthly* probability of unintended pregnancy outcomes
p_m_pill_birth <- RateToProb(rate_m_birth, 1/12)
p_m_pill_inducedabortion <- RateToProb(rate_m_inducedabortion, 1/12)
p_m_pill_spontaneousbaortion <- RateToProb(rate_m_spontaneousabortion, 1/12)
p_m_pill_ectopicpregnancy <- RateToProb(rate_m_ectopicpregnancy, 1/12)

# Unintended pregnancy outcomes by Strategy 
# These outcomes are probability of having an unintended pregnancy, given that you were on 
# one of the three strategies (do not sum to 1)

## *Monthly* *probabilities*
## Unintended pregnancy outcomes Strategy 1 - Pill only group
p_pill_birth <- p_m_pill_up * p_m_pill_birth
p_pill_inducedabortion <- p_m_pill_up * p_m_pill_inducedabortion
p_pill_spontaneousbaortion <- p_m_pill_up * p_m_pill_spontaneousbaortion
p_pill_ectopicpregnancy <- p_m_pill_up * p_m_pill_ectopicpregnancy


## Unintended pregnancy outcomes Strategy 2 - Pill + enzyme inducer group
p_pill_inducer <- p_m_pill_inducer_up * p_m_pill_birth
p_pill_inducerinducedabortion <- p_m_pill_inducer_up * p_m_pill_inducedabortion
p_pill_inducerspontaneousbaortion <- p_m_pill_inducer_up * p_m_pill_spontaneousbaortion
p_pill_inducerectopicpregnancy <- p_m_pill_inducer_up * p_m_pill_ectopicpregnancy


## Unintended pregnancy outcomes Strategy 3 - Pill + enzyme inhibitor group
p_pill_inhibitor_birth <- p_m_pill_inhibitor_up * p_m_pill_birth
p_pill_inhibitor_inducedabortion <- p_m_pill_inhibitor_up * p_m_pill_inducedabortion
p_pill_inhibitor_spontaneousbaortion <- p_m_pill_inhibitor_up * p_m_pill_spontaneousbaortion
p_pill_inhibitor_ectopicpregnancy <- p_m_pill_inhibitor_up * p_m_pill_ectopicpregnancy

## *Yearly* *Probability* of adverse event outcomes
p_vte <- 0.001
p_mi <- 0.000205
p_stroke <- 0.00037026

## *Yearly* *rate* of adverse event outcomes
r_vte <- ProbToRate(p_vte)
r_mi <- ProbToRate(p_mi)
r_stroke <- ProbToRate(p_stroke)

## *Monthly* *probabilities* of adverse events
p_m_vte <- RateToProb(r_vte, 1/12)
p_m_mi <- RateToProb(r_mi, 1/12)
p_m_stroke <- RateToProb(r_stroke, 1/12)


p_imb <- 0.005

## Probability of adverse event outcomes by strategy

## Probability of adverse event outcomes Strategy 1 - Pill
p_vte_pill <- p_m_vte
p_mi_pill <- p_m_mi
p_stroke_pill <- p_m_stroke
p_imb_pill <- p_imb

## Probability of adverse event outcomes Strategy 2 - Pill + enzyme inducer
p_vte_pill_inducer <- p_m_vte
p_mi_pill_inducer <- p_m_mi
p_stroke_pill_inducer <- p_m_stroke
p_imb_pill_inducer <- 0.07

## Probability of adverse event outcomes Strategy 3 - Pill + enzyme inhibitor
p_vte_pill_inhibitor <- p_m_vte + 0.2 * p_m_vte
p_mi_pill_inhibitor <- p_m_mi
p_stroke_pill_inhibitor <- p_m_stroke
p_imb_pill_inhibitor <- p_imb


# Costs of initial state for three strategies

c_pill  <- 680/12          # cost of remaining in pill only group
c_pill_inducer <- 340/12   # cost of remaining in pill + enzyme inducer group (conservative)
c_pill_inhibitor <- 340/12  # cost of remaining in pill + enzyme inhibitor group (conservative)

# Cost of birth outcomes 
c_birth  <- 13252              # cost of remaining one cycle sick
c_inducedabortion  <- 931      # cost of remaining one cycle dead
c_spontaneousabortion <- 1141  # cost of spontaneous abortion
c_ectopicpregnancy <- 6174     # cost of ectopic pregnancy

# cost of adverse event outcomes
c_vte <- 14305                # cost of venous thromboembolism
c_mi <- 23528                 # cost of myocardial infarction
c_stroke <- 28979             # cost of stroke

c_imb <- 0                   # cost of intermenstrual bleeding (conservative)
 

# Utility values 

## Utility of initial state for three strategies 
u_pill  <-  1                 # Assuming perfect health on pill alone                   
u_pill_inducer <- 0.95        # Lowered utility because of condition that requires drug (epilepsy, HIV etc)
u_pill_inhibitor <- 0.95      # Lowered utility because of condition that requires drug (epilepsy, HIV etc)

  
u_birth  <-  0.9064              # Utility of birth     
u_inducedabortion  <- 0.9615     # Utility of abortion
u_spontaneousabortion <- 0.9423  # Utility of spontaneous abortion
u_ectopicpregnancy <- 0.91667    # Utility of ectopic pregnancy
  
u_vte <- 0.81                   # Utility of VTE
u_mi <- 0.67                    # Utility of MI
u_stroke <- 0.33                # Utility of Stroke
u_imb <- 0.78                   # Utility of Inter menstrual bleeding

#d_e <- d_c <- 0.03                      # equal discount of costs and QALYs by 3%

# calculate discount weights for costs for each cycle based on discount rate d_c
#v_dwc <- 1 / (1 + d_e) ^ (0:n_t) 
# calculate discount weights for effectiveness for each cycle based on discount rate d_e
#v_dwe <- 1 / (1 + d_c) ^ (0:n_t) 
```

## ################################################################################################
## ################################################################################################
## ################################################################################################

### STRATEGY 1 PILL ONLY

## Draw the state-transition cohort model for the strategy 1 - "Pill only"
```{r}
m_P_diag_pill <- matrix(0, nrow = n_states, ncol = n_states, dimnames = list(v_n, v_n))
m_P_diag_pill["Initial", "Initial" ]  = "" 
m_P_diag_pill["Initial", "Birth" ]  = ""
m_P_diag_pill["Initial", "Induced abortion" ]  = ""
m_P_diag_pill["Initial", "Spontanous abortion" ]  = ""
m_P_diag_pill["Initial", "Ectopic pregnancy" ]  = ""
m_P_diag_pill["Initial", "VTE" ]  = ""
m_P_diag_pill["Initial", "MI" ]  = ""
m_P_diag_pill["Initial", "Stroke" ]  = ""
m_P_diag_pill["Initial", "IMB" ]  = ""
m_P_diag_pill["IMB", "IMB quitters"]  = ""
m_P_diag_pill["IMB", "IMB continuers"]  = ""
m_P_diag_pill["IMB continuers", "Initial"]  = ""

layout.fig <- c(1,4,4, 2)
st_tr_diag_pill <- plotmat(t(m_P_diag_pill), t(layout.fig), self.cex = 0.5, curve = 0, arr.pos = 0.8,  
        latex = T, arr.type = "curved", relsize = 0.85, box.prop = 0.8, 
        cex = 0.8, box.cex = 0.7, lwd = 1)
```

# 04 Define and initialize matrices and vectors

## 04.1 Cohort trace for the strategy 1 - "Pill only"

```{r}
# create the cohort trace
m_M_pill <- matrix(NA, 
              nrow = n_t + 1 ,  # create Markov trace (n.t + 1 because R doesn't 
                                # understand Cycle 0)
              ncol = n_states, 
              dimnames = list(0:n_t, v_n))

m_M_pill[1, ] <- c(1,0,0,0,0,0,0,0,0,0,0)          # initialize first cycle of Markov trace
```

## 04.2 Transition probability matrix for the strategy 1 - "Pill only"

```{r}
# create the transition probability matrix
m_P_pill  <- matrix(0,
               nrow = n_states, ncol = n_states,
               dimnames = list(v_n, v_n)) # name the columns and rows of the transition 
                                          # probability matrix
m_P_pill
```

Fill in the transition probability matrix for the strategy 1 - "Pill only":

```{r}
# from Healthy
m_P_pill["Initial", "Initial" ] <- 1 - p_m_pill_birth - p_m_pill_inducedabortion - p_m_pill_spontaneousbaortion - p_m_pill_ectopicpregnancy - p_vte_pill - p_mi_pill - p_stroke_pill - p_imb_pill 
m_P_pill["Initial", "Birth" ]    <- p_m_pill_birth 
m_P_pill["Initial", "Induced abortion" ]      <- p_m_pill_inducedabortion 
m_P_pill["Initial", "Spontanous abortion" ]  <- p_m_pill_spontaneousbaortion 
m_P_pill["Initial", "Ectopic pregnancy" ]   <- p_m_pill_ectopicpregnancy 
m_P_pill["Initial", "VTE" ] <-  p_vte_pill
m_P_pill["Initial", "MI" ]  <- p_mi_pill
m_P_pill["Initial", "Stroke" ] <-  p_stroke_pill 
m_P_pill["Initial", "IMB" ]  <- p_imb_pill 
m_P_pill["Birth", "Birth" ] <- 1
m_P_pill["Induced abortion", "Induced abortion" ] <- 1
m_P_pill["Spontanous abortion", "Spontanous abortion" ] <- 1
m_P_pill["Ectopic pregnancy", "Ectopic pregnancy" ] <- 1
m_P_pill["VTE", "VTE" ] <- 1
m_P_pill["MI", "MI" ] <- 1
m_P_pill["Stroke", "Stroke" ] <- 1
m_P_pill["IMB", "IMB quitters" ] <- 0.2
m_P_pill["IMB", "IMB continuers"] <- 0.8
m_P_pill["IMB quitters", "IMB quitters"] <- 1
m_P_pill["IMB continuers", "Initial"] <- 1


# check rows add up to 1
rowSums(m_P_pill)
```

# 05 Run Markov model for the strategy 1 - "Pill only"

```{r}
for (t in 1:n_t){                   # loop through the number of cycles
  m_M_pill[t + 1, ] <- m_M_pill[t, ] %*% m_P_pill  # estimate the state vector for the next cycle (t + 1)
}

```

# 06 Compute and Plot Outcomes for the strategy 1 - "Pill only"

## 06.1 Cohort trace for the strategy 1 - "Pill only"

```{r}
df_M_pill <- as.data.frame(m_M_pill) %>%
  tibble::rowid_to_column("cycle") %>%
  tidyr::gather(state, probability, -1)

ggplot(df_M_pill, aes(cycle, probability, color = state)) +
  geom_line()

```

## 06.2 Total Probability for Unintended pregnancy and adverse event outcomes for the strategy 1 - "Pill only"

```{r}
v_up_pill <- rowSums(m_M_pill[, 2:5])           # alternative way of calculating the OS probability   

v_adverseevents_pill <- rowSums(m_M_pill[, 6:9])

```


# 07 Compute Cost-Effectiveness Outcomes for the strategy 1 - "Pill only"

## 07.1 Mean Costs and QALYs for the strategy 1 - "Pill only"

```{r}
# per cycle
# calculate expected costs by multiplying m_M with the cost vector for the different 
# health states   
v_tc_pill <- m_M_pill %*% c(c_pill, c_birth, c_inducedabortion, c_spontaneousabortion,
                  c_ectopicpregnancy, c_vte, c_mi, c_stroke, c_imb)  
# calculate expected QALYs by multiplying m_M with the utilities for the different 
# health states   
v_tu_pill <- m_M_pill %*% c(u_pill, u_birth, u_inducedabortion, u_spontaneousabortion,
                  u_ectopicpregnancy, u_vte, u_mi, u_stroke, u_imb)  
```

## 07.2 Discounted Mean Costs and QALYs for the strategy 1 - "Pill only"

```{r}
# Discount costs by multiplying the cost vector with discount weights (v_dw) 
# v_tc_d_pill <-  t(v_tc_pill) %*% v_dwc
# Discount QALYS by multiplying the QALYs vector with discount weights (v_dw)
# v_te_d_pill <-  t(v_tu_pill) %*% v_dwe
```

## 07.3 Results for the strategy 1 - "Pill only"

```{r}
results_pill <- data.frame( "Total Discounted Cost" = v_tc_d_pill, 
                   
                       "Total Discounted QALYs" = v_te_d_pill, 
                       check.names = F)
results_pill
```
